---
title: "Process Samples for GEO Upload"
author:
  name: Christof Seiler
  affiliation: Department of Statistics, Stanford University
output:
  BiocStyle::pdf_document:
    toc: TRUE
    fig_width: 5
    fig_height: 4
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Goal

Process samples that have no sample information to be included in the main analysis. We need to do this because GEO requires all samples to be processed.

# Prerequisites

Install necessary packages from bioconductor repository. Run this code only once to install packages.

```{r install_packages, warning=FALSE, message=FALSE}
pkgs_needed = c("oligo","limma","hta20transcriptcluster.db","pd.hta.2.0",
                "affycoretools","genefilter","MatchIt","optmatch","ggfortify",
                "magrittr","statmod","readr","dplyr","readxl","stringr",
                "tibble","ggrepel")
letsinstall = setdiff(pkgs_needed, installed.packages()) 
if (length(letsinstall) > 0) {
  source("http://bioconductor.org/biocLite.R")
  biocLite(letsinstall)
}
```

Load packages.

```{r load_packages, warning=FALSE, message=FALSE}
library("oligo")
library("limma")
library("hta20transcriptcluster.db")
library("affycoretools")
library("genefilter")
library("MatchIt")
library("ggfortify")
library("magrittr")
library("statmod")
library("readr")
library("dplyr")
library("readxl")
library("stringr")
library("tibble")
library("ggrepel")
```

# Import Data

Load Affymetrix CEL files. At this stage, Bioconductor will automatically download the necessary annotation packages and install them for us. Load samples with missing sample information.

```{r read_celfiles}
sample_table = tibble(
  sample_name = c("4.1_Unkown_Mock",
                  "18.2_Unkown_H1N1"),
  file_name = c("Nicholas Bayless_US 4.1_(HTA-2_0).CEL",
                "Nicholas Bayless_H1N1 18.2_(HTA-2_0).CEL"),
  treatment = c("Mock", "H1N1")
)
pd = as(as.data.frame(sample_table), "AnnotatedDataFrame")
rawData = read.celfiles(sample_table$file_name,
                        phenoData = pd,
                        sampleNames = sample_table$sample_name)
rawData
```

# Quality Control

MA plots on the first three samples.

```{r maplot, out.width="50%"}
MAplot(rawData, pairs=TRUE)
```

# Preprocessing

Background subtraction, normalization and summarization using median-polish.

```{r normalizing}
eset = rma(rawData)
```

Get rid of background probes and annotate using functions in `affycoretools` package.

```{r affycoretools}
dbGetQuery(db(pd.hta.2.0), "select * from type_dict;")
table(getMainProbes("pd.hta.2.0")$type)
eset = getMainProbes(eset)
```

Filter probes that we cannot map to symbols.

```{r filter_probes}
e2s = toTable(hta20transcriptclusterSYMBOL)
prob_ids = rownames(exprs(eset))
keep_ids = which(prob_ids %in% e2s$probe_id)
eset = ExpressionSet(assayData = exprs(eset)[keep_ids,],
                     phenoData = phenoData(eset),
                     experimentData = experimentData(eset),
                     annotation = annotation(eset))
```

Write processed expressions to file for GEO upload.

```{r geo_upload}
geo_exprs_rma = exprs(eset)
colnames(geo_exprs_rma) = pData(eset)$sample_name
geo_exprs_rma %<>% as_tibble(rownames = "ID_REF")
write_csv(geo_exprs_rma, path = paste0("not_analyzed_rma.csv"))
```

# Session Info {.unnumbered}

```{r session_info}
sessionInfo()
```
