---
title: "Oligo with Matching"
author: Christof Seiler
date: November, 2016
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Goal

1. Load and normalize data using ``oligo``
2. Differential analysis using ``limma``

## Install Packages

Install necessary packages from bioconductor repository. Run this code only once to install packages.

```{r install_packages, warning=FALSE, message=FALSE, eval=FALSE}
pkgs_needed = c("oligo","limma",
                "hta20transcriptcluster.db","pd.hta.2.0",
                "MatchIt",
                "ggfortify","magrittr","statmod","readr","dplyr","readxl","stringr")
letsinstall = setdiff(pkgs_needed, installed.packages()) 
if (length(letsinstall) > 0) {
  source("http://bioconductor.org/biocLite.R")
  biocLite(letsinstall)
}
```

Load packages.

```{r load_packages}
library("oligo")
library("limma")
library("hta20transcriptcluster.db")
library("MatchIt")
library("ggfortify")
library("magrittr")
library("statmod")
library("readr")
library("dplyr")
library("readxl")
library("stringr")
```

## Import Data

Read sample tables. Merge with sample information table.

```{r read_sample_table}
sample_table = read_csv("sample_table_from_word.csv")
new_matched = read_excel("MSS Case_Control Matches (Masked)_new_matched.xlsx")
sample_table %<>% dplyr::left_join(new_matched,by = "ptidno", suffix = c("", "2"))
sample_table %<>% dplyr::select(ptidno,
                                ptb_case,
                                gestage_pbmc_collected,
                                gestage_delivery,
                                sample_id = SampleID)
```

Find CEL files in current folder.

```{r find_cel}
file_names_cel = list.files("./",pattern = "CEL")
tb_file_name = lapply(sample_table$sample_id,function(id) {
  pattern = paste0("[ _]",id)
  name = file_names_cel[which(str_detect(file_names_cel,pattern))]
  if(length(name)==0) name = NA
  tibble(sample_id = id,file_name = name)
}) %>% bind_rows()
sample_table %<>% dplyr::left_join(tb_file_name,by = "sample_id")
sample_table %<>% na.omit
```

Match samples.

```{r match_samples}
sample_table$group = sample_table$ptb_case == "case"
mout = matchit(group ~ gestage_pbmc_collected, 
               data = sample_table, 
               method = "nearest", 
               ratio = 1)
summary(mout)
sample_table$pair = NA
A = rownames(mout$match.matrix) %>% as.integer
B = mout$match.matrix %>% as.integer
for(i in 1:nrow(mout$match.matrix)) {
  sample_table$pair[A[i]] = i
  sample_table$pair[B[i]] = i
}
sample_table %<>% dplyr::select(-group)
#sample_table %<>% na.omit
#write_csv(sample_table,path = "sample_table_matched.csv")
```

Then load Affymetrix CEL files. At this stage, Bioconductor will automatically download the necessary annotation packages and install them for us. 

```{r read_celfiles}
pd = as(as.data.frame(sample_table), "AnnotatedDataFrame")
rawData = read.celfiles(sample_table$file_name,
                        phenoData = pd,
                        sampleNames = sample_table$sample_id)
rawData
```

## Quality Control

MA plots on the first three samples.

```{r maplot}
MAplot(rawData[, 1:3], pairs=TRUE)
```

## Preprocessing

Background subtraction, normalization and summarization using median-polish.

```{r normalizing}
eset = rma(rawData)
class(eset)
show(eset)
exprs(eset)[1:10, 1:2]
save(eset,file = "eset.Rdata")
```

PCA plot of normalized expressions.

```{r pca_plots}
res_pca = prcomp(t(exprs(eset)), scale. = FALSE)
screeplot(res_pca)
sample_table_annotated = eset@phenoData@data
sample_table_annotated$pair %<>% as.factor
asp_ratio = res_pca$sdev[2]^2/res_pca$sdev[1]^2
autoplot(res_pca, 
         data = sample_table_annotated, 
         shape = "ptb_case", 
         colour = "gestage_pbmc_collected",
         size = 3, 
         asp = asp_ratio)
autoplot(res_pca, 
         data = sample_table_annotated, 
         shape = "ptb_case", 
         colour = "gestage_delivery",
         size = 3,
         asp = asp_ratio)
rownames(sample_table_annotated) = paste0(sample_table_annotated$pair,
                                          "_",
                                          sample_table_annotated$sample_id)
autoplot(res_pca, 
         data = sample_table_annotated, 
         shape = FALSE, 
         colour = "pair",
         label = TRUE,
         asp = asp_ratio)
```

## Differential Expression Analyses

Use ``limma`` for linear models to assess difference in expression. Paired analysis as described in Section 9.4.1 on page 42 in the [limma vignette](https://www.bioconductor.org/packages/devel/bioc/vignettes/limma/inst/doc/usersguide.pdf).

```{r paired_samples}
targets = eset@phenoData@data
pair = factor(targets$pair)
treat = factor(targets$ptb_case, levels = c("control","case"))
design = model.matrix(~ pair + treat)
fit = lmFit(eset, design)
fit = eBayes(fit)
```

The topTable command provides us a way of ranking genes for further evaluation. In the case below, we adjust for multiple testing by FDR.

```{r}
gene_table = topTable(fit, coef = "treatcase", adjust = "BH", number = nrow(fit))
hist(gene_table$P.Value,breaks = 20)
sum(gene_table$adj.P.Val < 0.05)
```

Volcano plots for quality control.

```{r}
volcanoplot(fit, coef = "treatcase", highlight = 10)
```

Map between manufacturer identifiers and gene symbols.

```{r}
e2s = toTable(hta20transcriptclusterSYMBOL)
map_gene_symbol = function(gene_table) {
  prob_ids = rownames(gene_table)
  symbol = sapply(prob_ids,function(prob_id) {
    matching_symbol = e2s$symbol[prob_id==e2s$probe_id]
    if(length(matching_symbol)==0) matching_symbol = "No_Symbol_Found"
    matching_symbol
  }) %>% unlist
  gene_table = cbind(gene_table,symbol=symbol,stringsAsFactors=FALSE)
  print(head(gene_table))
  gene_table  
}
gene_table = map_gene_symbol(gene_table)
print(head(gene_table, n = 10))
```

## Export Results

Write to text file.

```{r}
write_csv(gene_table,path = "gene_table_results.csv")
```

## Session Info

```{r session_info}
sessionInfo()
```
