---
title: "Validation of RSG13"
author:
  name: Christof Seiler
  affiliation: Department of Statistics, Stanford University
output:
  BiocStyle::html_document2:
    toc_float: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Goal

Compare expression of RGS13 in related pregnency studies that are publicatly available. Use the following studies: 

1. [Whole blood gene expression profile associated with spontaneous preterm birth](https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE46510)
2. [Maternal Whole Blood Gene Expression at 18 and 28 weeks of Gestation Associated with Spontaneous Preterm Birth in Asymptomatic Women](https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE59491)
3. [Unique inflammatory transcriptome profiles at the maternal fetal interface and onset of human preterm and term birth](https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE73685)

# Prerequisites

Install necessary packages from bioconductor repository. Run this code only once to install packages.

```{r install_packages, warning=FALSE, message=FALSE}
pkgs_needed = c("hta20transcriptcluster.db",
                "hgu133plus2.db",
                "hugene21sttranscriptcluster.db",
                "hugene10sttranscriptcluster.db",
                "magrittr",
                "dplyr")
letsinstall = setdiff(pkgs_needed, installed.packages()) 
if (length(letsinstall) > 0) {
  source("http://bioconductor.org/biocLite.R")
  biocLite(letsinstall)
}
```

Load packages.

```{r load_packages, warning=FALSE, message=FALSE}
library("hta20transcriptcluster.db")
library("hgu133plus2.db")
library("hugene21sttranscriptcluster.db")
library("hugene10sttranscriptcluster.db")
library("magrittr")
library("dplyr")
```

# Check RGS13

Check if RGS13 is available on microarray.

```{r check_rgs13}
e2s_our = toTable(hta20transcriptclusterSYMBOL)
e2s_study1 = toTable(hgu133plus2SYMBOL)
e2s_study2 = toTable(hugene21sttranscriptclusterSYMBOL)
e2s_study3 = toTable(hugene10sttranscriptclusterSYMBOL)
e2s_our %>% dplyr::filter(symbol == "RGS13")
e2s_study1 %>% dplyr::filter(symbol == "RGS13")
e2s_study2 %>% dplyr::filter(symbol == "RGS13")
e2s_study3 %>% dplyr::filter(symbol == "RGS13")
```

# Session Info {.unnumbered}

```{r session_info}
sessionInfo()
```
