---
title: "Follow-Up qPCR Experiments"
author: Christof Seiler
date: April, 2018
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Goal

Analyze data from qPCR experiments.

## Install Packages

Install necessary packages from bioconductor repository. Run this code only once to install packages.

```{r install_packages, warning=FALSE, message=FALSE}
pkgs_needed = c("ggplot2","magrittr","readr","dplyr","readxl","stringr","tidyr","Rfit")
letsinstall = setdiff(pkgs_needed, installed.packages()) 
if (length(letsinstall) > 0) {
  source("http://bioconductor.org/biocLite.R")
  biocLite(letsinstall)
}
```

Load packages.

```{r load_packages, warning=FALSE, message=FALSE}
library("magrittr")
library("ggplot2")
library("readr")
library("dplyr")
library("readxl")
library("stringr")
library("tidyr")
library("Rfit")
```

## Import Data

Read, merge, and tidy sample tables.

```{r read_sample_table}
new_matched = read_excel("MSS Case_Control Matches (Masked)_new_matched.xlsx")
new_matched %<>% mutate(time_to_delivery_calc = gestage_delivery - gestage_enroll)
qpcr_data = read_excel("PTB Raw qPCR data.xlsx")
qpcr_data$SampleID %<>% as.character
qpcr_data %<>% dplyr::left_join(new_matched,by = "SampleID", suffix = c("", "2"))
qpcr_data %<>% dplyr::select(sample_id = SampleID,
                             ptb_case,
                             time_to_delivery = time_to_delivery_calc,
                             gestage_delivery,
                             RGS13_baseline,
                             IFNL_baseline,
                             IFNL_flu)
qpcr_data %<>% gather(RGS13_baseline, IFNL_baseline, IFNL_flu, 
                      key = "key", value = "expr")
qpcr_data %<>% mutate(treatment = if_else(condition = str_detect(qpcr_data$key,"baseline"),
                                          true = "mock",
                                          false = "H1N1"))
qpcr_data %<>% mutate(gene_name = sapply(strsplit(qpcr_data$key,split = "_"),function(str) str[[1]]))
qpcr_data %<>% select(-key)
qpcr_data$treatment = factor(qpcr_data$treatment,levels = c("mock","H1N1"))
qpcr_data
```

## Fit Model

Subset to RGS13.

```{r subset_RGS13}
qpcr_data_RGS13 = qpcr_data %>% dplyr::filter(gene_name == "RGS13")
```

Plotting to see possible outliers.

```{r plotting_RGS13,fig.width=4,fig.height=4}
ggplot(qpcr_data_RGS13,aes(expr)) + 
  geom_histogram(bins = 30) + 
  facet_wrap(~ptb_case) +
  ggtitle("RGS13")
ggplot(qpcr_data_RGS13,aes(x = time_to_delivery, y = expr, color = treatment)) +
  geom_smooth(method = lm, se = FALSE) +
  geom_point() +
  geom_text(aes(label = ifelse(expr > quantile(expr, probs = 0.95), sample_id,"")),
            hjust = 0, vjust = 0) +
  ggtitle("RGS13")
ggsave(filename = "RGS13.png",width = 4,height = 4,dpi = 600)
ggplot(qpcr_data_RGS13,aes(x = gestage_delivery, y = expr, color = treatment)) +
  geom_smooth(method = lm, se = FALSE) +
  geom_point() +
    geom_text(aes(label = ifelse(expr > quantile(expr, probs = 0.95), sample_id,"")),
            hjust = 0, vjust = 0) +
  ggtitle("RGS13")
```

Fit model for `RGS13`.

```{r fit_model_RGS13}
lm_fit = lm(expr ~ time_to_delivery + gestage_delivery, qpcr_data_RGS13)
summary(lm_fit)
```

Fit robust model using rank-based estimates of regression coefficients.

```{r fit_model_RGS13_rank}
lm_fit = rfit(expr ~ time_to_delivery + gestage_delivery, qpcr_data_RGS13)
summary(lm_fit)
```

Remove potential outlier and refit model to see if we still obtain a similar result.

```{r fit_model_RGS13_outlier}
qpcr_data_RGS13 %<>% dplyr::filter(sample_id != "16.2")
lm_fit = lm(expr ~ time_to_delivery + gestage_delivery, qpcr_data_RGS13)
summary(lm_fit)
```

Subset to `IFNL`.

```{r subset_IFNL}
qpcr_data_IFNL = qpcr_data %>% dplyr::filter(gene_name == "IFNL")
```

Plotting to see possible outliers.

```{r plotting_IFNL,fig.width=4,fig.height=4}
ggplot(qpcr_data_IFNL,aes(expr)) + 
  geom_histogram(bins = 30) + 
  facet_wrap(~ptb_case) +
  ggtitle("IFNL")
ggplot(qpcr_data_IFNL,aes(x = time_to_delivery, y = expr, color = treatment)) +
  geom_smooth(method = lm, se = FALSE) +
  geom_point() +
  geom_text(aes(label = ifelse(expr > quantile(expr, probs = 0.95), sample_id,"")),
            hjust = 0, vjust = 0) +
  ggtitle("IFNL")
ggsave(filename = "IFNL.png",width = 4,height = 4,dpi = 600)
ggplot(qpcr_data_IFNL,aes(x = gestage_delivery, y = expr, color = treatment)) +
  geom_smooth(method = lm, se = FALSE) +
  geom_point() +
    geom_text(aes(label = ifelse(expr > quantile(expr, probs = 0.95), sample_id,"")),
            hjust = 0, vjust = 0) +
  ggtitle("IFNL")
```

Fit model for `IFNL`.

```{r fit_model_IFNL}
lm_fit = lm(expr ~ gestage_delivery + 
               treatment + 
               time_to_delivery + 
               time_to_delivery:treatment +
               gestage_delivery:treatment, 
            data = qpcr_data_IFNL)
summary(lm_fit)
```

Fit robust model using rank-based estimates of regression coefficients.

```{r fit_model_rank}
lm_fit = rfit(expr ~ gestage_delivery + 
               treatment + 
               time_to_delivery + 
               time_to_delivery:treatment +
               gestage_delivery:treatment, 
            data = qpcr_data_IFNL)
summary(lm_fit)
```

Refit linear model after removing possible outlier.

```{r fit_model_IFNL_outlier}
qpcr_data_IFNL %<>% dplyr::filter(sample_id != "1.1")
lm_fit = lm(expr ~ gestage_delivery + 
               treatment + 
               time_to_delivery + 
               time_to_delivery:treatment +
               gestage_delivery:treatment, 
            data = qpcr_data_IFNL)
summary(lm_fit)
```

## Session Info

```{r session_info}
sessionInfo()
```
