---
title: "Time to Delivery on Microarray Data"
author: Christof Seiler
date: February, 2017
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Goal

1. Load and normalize data using ``oligo``
2. Differential analysis using ``limma``

## Install Packages

Install necessary packages from bioconductor repository. Run this code only once to install packages.

```{r install_packages, warning=FALSE, message=FALSE}
pkgs_needed = c("oligo","limma","ggfortify","magrittr","hta20transcriptcluster.db",
                "statmod","stringr")
letsinstall = setdiff(pkgs_needed, installed.packages()) 
if (length(letsinstall) > 0) {
  source("http://bioconductor.org/biocLite.R")
  biocLite(letsinstall)
}
```

Load packages.

```{r load_packages, warning=FALSE, message=FALSE}
library("oligo")
library("limma")
library("ggfortify")
library("magrittr")
library("hta20transcriptcluster.db")
library("statmod")
library("stringr")
```

## Import Data

Then load Affymetrix CEL files. At this stage, Bioconductor will automatically download the necessary annotation packages and install them for us. 

```{r}
sample_table = read.csv("sample_table.csv")
cel_filenames = paste0(as.character(sample_table$array_name),".CEL")
pd = as(sample_table, "AnnotatedDataFrame")
sample_names = paste(sample_table$sample_id,
                     sample_table$condition,
                     sample_table$treatment,sep = "_")
rawData = read.celfiles(cel_filenames,phenoData = pd,sampleNames = sample_names)
rawData
```

This is the downloaded annotation package:

```{r}
pd.hta.2.0
```

Let's have a first look at the data. These are the samples:

```{r}
slotNames(rawData)
sampleNames(rawData)
```

## Preprocessing

Background subtraction, normalization and summarization using median-polish.

```{r}
eset = rma(rawData)
class(eset)
show(eset)
exprs(eset)[1:10, 1:2]
save(eset,file = "eset.Rdata")
```

PCA plot of normalized expressions.

```{r}
res_pca = prcomp(t(exprs(eset)),scale. = FALSE)
screeplot(res_pca)
sample_table = eset@phenoData@data
autoplot(res_pca, data = sample_table, colour = 'treatment', shape = 'condition', size = 3)
sample_names = paste(sample_table$sample_id,
                     sample_table$treatment,sep = "_")
rownames(sample_table) = sample_names
autoplot(res_pca, data = sample_table, 
         shape = FALSE, label = TRUE, label.size = 3)
```

Add time to delivery variable.

```{r}
sample_table$time_to_delivery = sample_table$gestage_delivery - sample_table$gestage_enroll
```

Annotate PCA with experimental information.

```{r}
sample_table$run_batch = as.factor(sample_table$run_batch)
autoplot(res_pca, data = sample_table, colour = 'rin', size = 3)
autoplot(res_pca, data = sample_table, colour = 'run_day', size = 3)
autoplot(res_pca, data = sample_table, colour = 'run_batch', size = 3)
autoplot(res_pca, data = sample_table, colour = 'viable_cell_count', size = 3)
autoplot(res_pca, data = sample_table, colour = 'viability', size = 3)
autoplot(res_pca, data = sample_table, colour = 'gestage_delivery', size = 3)
autoplot(res_pca, data = sample_table, colour = 'gestage_enroll', size = 3)
autoplot(res_pca, data = sample_table, colour = 'time_to_delivery', size = 3)
```

## Differential Expression Analyses

Use ``limma`` for linear models to assess difference in expression. Define design matrix.

```{r}
sample_table$sample_id = factor(sample_table$sample_id)
sample_table$treatment = factor(sample_table$treatment,levels = c("Mock","H1N1"))
design = model.matrix(~ time_to_delivery * treatment + gestage_delivery * treatment,
                      sample_table)
colnames(design) = str_replace(string = colnames(design),pattern = ":",replacement = "_") %>%
  str_replace(string = .,pattern = "treatment",replacement = "")
colnames(design)
```

Estimate the correlation between measurements made on the same subject.

```{r}
corfit = duplicateCorrelation(eset, design,block = sample_table$sample_id)
```

Then this inter-subject correlation is input into the linear model fit.

```{r}
fit = lmFit(eset,design,block = sample_table$sample_id,correlation = corfit$consensus)
```

Then compute moderated t-tests.

```{r}
fit = eBayes(fit)
```

The topTable command provides us a way of ranking genes for further evaluation. In the case below, we adjust for multiple testing by FDR. Save results in a list of tables.

```{r}
coeffs = colnames(design)[-1]
gene_table_list = lapply(coeffs,function(coeff_name) {
  gene_table = topTable(fit, 
                        coef = coeff_name, 
                        adjust="BH", 
                        number = nrow(fit))
  hist(gene_table$P.Value,breaks = 100,main = coeff_name)
  gene_table
})
names(gene_table_list) = coeffs
```

Volcano plots for quality control.

```{r}
for(coeff_name in coeffs) {
  cat(coeff_name,"\n")
  volcanoplot(fit,coef = coeff_name,highlight = 10)
}
```

Map between manufacturer identifiers and gene symbols.

```{r}
e2s = toTable(hta20transcriptclusterSYMBOL)
map_gene_symbol = function(gene_table) {
  prob_ids = rownames(gene_table)
  symbol = sapply(prob_ids,function(prob_id) {
    matching_symbol = e2s$symbol[prob_id==e2s$probe_id]
    if(length(matching_symbol)==0) matching_symbol = "No_Symbol_Found"
    matching_symbol
  }) %>% unlist
  gene_table = cbind(gene_table,symbol=symbol,stringsAsFactors=FALSE)
  gene_table
}
gene_table_list = lapply(gene_table_list,map_gene_symbol)
```

Print genes that are below an FDR of 0.05.

```{r}
for(i in 1:length(gene_table_list)) {
  separator = "-----------"
  cat(separator,names(gene_table_list)[i],separator,"\n")
  gene_table_subset = subset(gene_table_list[[i]],adj.P.Val < 0.05)
  print(gene_table_subset[,c("logFC","adj.P.Val","symbol")])
  cat(separator,names(gene_table_list)[i],separator,"\n")
}
```

Write to text file.

```{r}
for(i in 1:length(gene_table_list)) {
  file_name_results = paste0(names(gene_table_list)[i],"_results.csv")
  cat("writting:",file_name_results,"\n")
  write.csv(gene_table_list[[i]],file = file_name_results)
}
```

## Plot Individual Probes

Plot the raw abundances for individual probes against explanatory variables. Run day effects?

```{r}
plot_probe = function(probe_name) {
  probe_id = which(rownames(exprs(eset)) == probe_name)
  df_probe = data.frame(probe = exprs(eset)[probe_id,],
                        sample_table)
  ggplot(df_probe, aes(x = time_to_delivery,y = probe,color = run_day)) + 
    geom_point() + 
    facet_wrap(~ treatment) +
    ggtitle(probe_name)
}
plot_probe("TC14000062.hg.1")
plot_probe("ERCCmix2step23")
plot_probe("TC11002981.hg.1")
```

Difference in slope.

```{r}
plot_probe = function(probe_name) {
  probe_id = which(rownames(exprs(eset)) == probe_name)
  df_probe = data.frame(probe = exprs(eset)[probe_id,],
                        sample_table)
  ggplot(df_probe, aes(x = time_to_delivery,y = probe)) + 
    geom_point() + 
    geom_smooth(method = "lm", se = FALSE) +
    facet_wrap(~ treatment) +
    ggtitle(probe_name)
}
plot_probe("TC01001623.hg.1")
```

Difference in mean between H1N1 and Mock.

```{r}
plot_probe("TC11001706.hg.1")
```

Difference in slope between H1N1 and Mock.

```{r}
plot_probe("TC19000540.hg.1")
```

## Session Info

```{r session_info}
sessionInfo()
```
