---
title: "Time to Delivery Analysis on Follow-Up qPCR Data"
author:
  name: Christof Seiler
  affiliation: Department of Statistics, Stanford University
output:
  BiocStyle::html_document:
    toc_float: true
params:
  treatment: "H1N1"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Goal

Analyze data from qPCR experiments.

# Prerequisites

Install necessary packages from bioconductor repository. Run this code only once to install packages.

```{r install_packages, warning=FALSE, message=FALSE}
pkgs_needed = c("ggplot2","magrittr","readr","dplyr",
                "readxl","stringr","tidyr","car",
                "interactions","ggthemes","cowplot",
                "MASS", "boot")
letsinstall = setdiff(pkgs_needed, installed.packages()) 
if (length(letsinstall) > 0) {
  source("http://bioconductor.org/biocLite.R")
  biocLite(letsinstall)
}
```

Load packages.

```{r load_packages, warning=FALSE, message=FALSE}
library("magrittr")
library("ggplot2")
library("readr")
library("dplyr")
library("readxl")
library("stringr")
library("tidyr")
library("MASS")
library("interactions")
library("ggthemes")
library("cowplot")
library("car")
library("boot")
theme_set(theme_few())
scale_colour_discrete = function(...) scale_colour_few()
if(params$treatment == "H1N1") {
  color_condition = "#FAA43A"
} else {
  color_condition = "#5DA5DA"
}
```

# Import Data

Read, merge, and tidy sample tables.

```{r read_sample_table}
sample_info_filename = "MSS Case_Control Matches (Masked)_new_matched.xlsx"
data_filename = "PTB Raw qPCR data.xlsx"
new_matched = read_excel(sample_info_filename)
new_matched %<>% mutate(time_to_delivery_calc = 
                          gestage_enroll - gestage_delivery)
qpcr_data = read_excel(data_filename)
qpcr_data$SampleID %<>% as.character
qpcr_data %<>% dplyr::left_join(new_matched,by = "SampleID", 
                                suffix = c("", "2"))
qpcr_data %<>% dplyr::select(sample_id = SampleID,
                             ptb_case,
                             time_to_delivery = time_to_delivery_calc,
                             gestage_delivery,
                             RGS13_baseline,
                             IFNL_baseline,
                             IFNL_flu)
qpcr_data %<>% gather(RGS13_baseline, IFNL_baseline, IFNL_flu, 
                      key = "key", value = "expr")
qpcr_data %<>% mutate(
  treatment = if_else(condition = str_detect(qpcr_data$key, "baseline"),
                      true = "mock",
                      false = "H1N1"))
qpcr_data %<>% mutate(
  gene_name = sapply(strsplit(qpcr_data$key,split = "_"),
                     function(str) str[[1]]))
qpcr_data %<>% dplyr::select(-key)
qpcr_data$treatment = factor(qpcr_data$treatment,levels = c("mock","H1N1"))
qpcr_data %<>% rename(ptb = ptb_case)
qpcr_data
```

# Fit Model

## RGS13

Subset to RGS13.

```{r subset_RGS13}
qpcr_data_RGS13 = qpcr_data %>% dplyr::filter(gene_name == "RGS13")
table(qpcr_data_RGS13$gene_name)
```

Plotting to see possible outliers.

```{r plotting_RGS13,fig.width=4,fig.height=4}
ggplot(qpcr_data_RGS13,aes(expr)) + 
  geom_histogram(bins = 30) + 
  facet_wrap(~ptb) +
  ggtitle("RGS13")
ggplot(qpcr_data_RGS13,aes(x = time_to_delivery, y = expr, 
                           color = treatment)) +
  geom_smooth(method = lm, se = FALSE) +
  geom_point() +
  geom_text(aes(label = ifelse(expr > quantile(expr, probs = 0.95), 
                               sample_id,"")), hjust = 0, vjust = 0) +
  ggtitle("RGS13")
ggplot(qpcr_data_RGS13,aes(x = gestage_delivery, y = expr, 
                           color = treatment)) +
  geom_smooth(method = lm, se = FALSE) +
  geom_point() +
  geom_text(aes(label = ifelse(expr > quantile(expr, probs = 0.95), 
                               sample_id,"")), hjust = 0, vjust = 0) +
  ggtitle("RGS13")
```

Fit model for `RGS13`.

```{r fit_model_RGS13}
lm_fit = lm(expr ~ time_to_delivery + gestage_delivery, qpcr_data_RGS13)
summary(lm_fit)
```

Fit robust model using rank-based estimates of regression coefficients.

```{r fit_model_RGS13_rank}
lm_fit = rlm(expr ~ time_to_delivery + gestage_delivery, qpcr_data_RGS13)
summary(lm_fit)
```

Remove potential outlier and refit model to see if we still obtain a similar result.

```{r fit_model_RGS13_outlier}
qpcr_data_RGS13 %<>% dplyr::filter(sample_id != "16.2")
lm_fit = lm(expr ~ time_to_delivery + gestage_delivery, qpcr_data_RGS13)
summary(lm_fit)
```

Plot of fitted intercept and slope with `gestage_delivery` fixed at its mean value.

```{r fit_model_RGS13_outlier_plot, fig.width=4,fig.height=4}
coefs = coef(lm_fit)
gmine = ggplot(qpcr_data_RGS13, 
               aes(x = time_to_delivery, y = expr, shape = ptb, color = treatment)) + 
  geom_point(size = 2) + 
  scale_colour_manual(values = color_condition)

mean_gestage = mean(qpcr_data_RGS13$gestage_delivery)

gmine + geom_abline(intercept = coefs["(Intercept)"] + 
                      mean_gestage*coefs["gestage_delivery"],
                    slope = coefs["time_to_delivery"],
                    color = color_condition, size = 1) + ggtitle("RGS13")
ggsave(filename = "time_to_delivery_RGS13.png",
       width = 4,height = 3,dpi = 600)
```

## IFNL

Subset to `IFNL`.

```{r subset_IFNL}
qpcr_data_IFNL = qpcr_data %>% 
  dplyr::filter(gene_name == "IFNL") %>%
  dplyr::filter(treatment == params$treatment)
```

Plotting to see possible outliers.

```{r plotting_IFNL,fig.width=4,fig.height=4}
ggplot(qpcr_data_IFNL,aes(expr)) + 
  geom_histogram(bins = 30) + 
  facet_wrap(~ptb) +
  ggtitle("IFNL")
ggplot(qpcr_data_IFNL,aes(x = time_to_delivery, y = expr, 
                          color = treatment)) +
  geom_smooth(method = lm, se = FALSE) +
  geom_point() +
  geom_text(aes(label = ifelse(expr > quantile(expr, probs = 0.95), 
                               sample_id,"")), hjust = 0, vjust = 0) +
  ggtitle("IFNL")
ggplot(qpcr_data_IFNL,aes(x = gestage_delivery, y = expr, 
                          color = treatment)) +
  geom_smooth(method = lm, se = FALSE) +
  geom_point() +
  geom_text(aes(label = ifelse(expr > quantile(expr, probs = 0.95), 
                               sample_id,"")), hjust = 0, vjust = 0) +
  ggtitle("IFNL")
```

Fit model for `IFNL`.

```{r fit_model_IFNL}
lm_fit = lm(expr ~ time_to_delivery + gestage_delivery, data = qpcr_data_IFNL)
summary(lm_fit)
```

Fit robust linear model fit and bootstrap resampling for confidence intervals and p-values.

```{r fit_model_rank, fig.height=8}
lm_fit_robust = rlm(expr ~ time_to_delivery + gestage_delivery, data = qpcr_data_IFNL, maxit = 200)
summary(lm_fit_robust)
# bootstrap confidence interval and p-value
set.seed(0xdada)
lm_fit_robust_boot = Boot(lm_fit_robust, R = 1999)
summary(lm_fit_robust_boot)
hist(lm_fit_robust_boot, legend="separate")
Confint(lm_fit_robust_boot, level = 0.95, type = "perc")
Confint(lm_fit_robust_boot, level = 0.95, type = "bca")
# p-value for stimulation x time to delivery interaction term
R = 1999
t_obsv = lm_fit_robust_boot$t0["time_to_delivery"]
t_boot = lm_fit_robust_boot$t[,"time_to_delivery"]-t_obsv
pvalue = (1 + sum(abs(t_boot) > abs(t_obsv))) / (R + 1)
pvalue
```

Refit linear model after removing possible outlier.

```{r fit_model_IFNL_outlier}
qpcr_data_IFNL_with_outlier = qpcr_data_IFNL
lm_fit = lm(expr ~ time_to_delivery + gestage_delivery, 
            data = qpcr_data_IFNL %>% 
              dplyr::filter(sample_id != "1.1"))
summary(lm_fit)
```

Residual plots.

```{r residual_plots, fig.width=4, fig.height=4}
tb_subset = qpcr_data_IFNL
rlm_delivery = rlm(expr ~ gestage_delivery, data = tb_subset)
tb_subset %<>% mutate(residual_expr = rlm_delivery$residuals)
rlm_delivery = rlm(residual_expr ~ time_to_delivery, data = tb_subset)
a = rlm_delivery$coefficients[[1]]
b = rlm_delivery$coefficients[[2]]
tb_subset %<>% mutate(y_hat = a + b*tb_subset$time_to_delivery)
gmine_with_outlier = ggplot(tb_subset, aes(time_to_delivery, residual_expr, color = treatment)) +
  geom_abline(intercept = a, slope = b, color = color_condition, size = 1) +
  geom_point() + 
  scale_colour_manual(values = color_condition) +
  ggtitle("IFNL (robust fit)") +
  xlab("time to delivery") +
  ylab("residual expression") +
  theme(legend.position = "bottom")
gmine_with_outlier
ggsave(gmine_with_outlier, 
       filename = paste0("robust_fit_IFNL_", params$treatment, ".png"),
       width = 3, height = 4, dpi = 600)
```

# Session Info {.unnumbered}

```{r session_info}
sessionInfo()
```
