---
title: "Export Gene Table to Word"
author: "Christof Seiler"
date: "January 2020"
output: word_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Reviewer Request

A reviewer requested the full table with all 170 genes at FDR = 0.1.

From [limma's user guide](https://www.bioconductor.org/packages/release/bioc/vignettes/limma/inst/doc/usersguide.pdf):

* `logFC`: log2-expression level
* `AveExpr`: average log2-expression level for that gene across all the arrays and channels in the experiment
* `t`: moderated t-statistic
* `P.Value`: $p$-value
* `adj.P.Value`:  $p$-value adjusted for multiple testing with the Benjamini and Hochberg method, also called $q$-values

Meaning of $q$-values: If all genes with $q$-value below a threshold, say 0.05, are selected as differentially expressed, then the expected proportion of false discoveries in the selected group is controlled to be less than the threshold value, in this case 5%.

```{r prepare_table, message=FALSE, warning=FALSE}
library("readr")
library("dplyr")
library("magrittr")
table = read_csv("time_to_delivery_time_to_delivery_H1N1.csv")
table %<>% filter(adj.P.Val < 0.1)
table %<>% arrange(P.Value)
table %<>% select(symbol, logFC, AveExpr, P.Value, adj.P.Val)
table %<>% mutate(
  AveExpr = format(AveExpr, digits = 2),
  logFC = format(logFC, digits = 1), 
  P.Value = format(P.Value, digits = 1, scientific = TRUE),
  adj.P.Val = format(adj.P.Val, digits = 2)
)
table %<>% dplyr::rename(
  `Gene symbol` = symbol,
  `Average expression (log2)` = AveExpr,
  `Coefficient (log2)` = logFC, 
  `Unadjusted p-value` = P.Value,
  `Adjusted p-value` = adj.P.Val
  )
```

```{r write_table}
knitr::kable(
  table, align = "lrrrr", 
  caption = "Table 2. Differentially expressed genes in stimulated samples for time to deliver term. We included the following covariates in the linear model: intercept, time to delivery, and gestational age at birth. In the table, we show the top 170 genes for the time to delivery term. See Figure 2 for a full network analysis on all genes. The full table is in the supplementary material."
)
```
