---
title: "Separate Mock/H1N1 Time to Delivery Analysis on Microarray Data"
author:
  name: Christof Seiler
  affiliation: Department of Statistics, Stanford University
output:
  BiocStyle::html_document2:
    toc_float: true
params:
  treatment: "H1N1"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Goal

1. Load and normalize data using ``oligo``
2. Differential analysis using ``limma``

# Prerequisites

Install necessary packages from bioconductor repository. Run this code only once to install packages.

```{r install_packages, warning=FALSE, message=FALSE}
pkgs_needed = c("oligo","limma","affycoretools","genefilter",
                "hta20transcriptcluster.db","ggfortify","magrittr",
                "statmod","stringr","tibble","dplyr",
                "STRINGdb","BioNet","DLBCL","org.Hs.eg.db","KEGGREST",
                "igraph","intergraph","ggnetwork","ggthemes")
letsinstall = setdiff(pkgs_needed, installed.packages()) 
if (length(letsinstall) > 0) {
  source("http://bioconductor.org/biocLite.R")
  biocLite(letsinstall)
}
```

Load packages.

```{r load_packages, warning=FALSE, message=FALSE}
library("oligo")
library("limma")
library("affycoretools")
library("genefilter")
library("hta20transcriptcluster.db")
library("ggfortify")
library("magrittr")
library("statmod")
library("stringr")
library("tibble")
library("dplyr")
library("STRINGdb")
library("BioNet")
library("DLBCL")
library("org.Hs.eg.db")
library("KEGGREST")
library("igraph")
library("intergraph")
library("ggnetwork")
library("ggthemes")
library("sna")
library("statnet.common")
library("network")
theme_set(theme_few())
scale_colour_discrete = function(...) scale_colour_few()
```

# Import Data

Then load Affymetrix CEL files. At this stage, Bioconductor will automatically download the necessary annotation packages and install them for us. Add time to delivery variable.

```{r import_data}
sample_table = read.csv("sample_table.csv")
params$treatment
sample_table %<>% filter(treatment == params$treatment)
sample_table$time_to_delivery = sample_table$gestage_delivery - 
  sample_table$gestage_enroll
sample_names = paste(sample_table$sample_id,
                     sample_table$condition,
                     sample_table$treatment,sep = "_")
pd = as(sample_table, "AnnotatedDataFrame")
cel_filenames = paste0(as.character(sample_table$array_name),".CEL")
rawData = read.celfiles(cel_filenames, phenoData = pd, 
                        sampleNames = sample_names)
rawData
```

# Preprocessing

Background subtraction, normalization and summarization using median-polish.

```{r rma}
eset = rma(rawData)
```

Get rid of background probes and annotate using functions in `affycoretools` package.

```{r affycoretools}
dbGetQuery(db(pd.hta.2.0), "select * from type_dict;")
table(getMainProbes("pd.hta.2.0")$type)
eset = getMainProbes(eset)
```

Filter probes that we cannot map to symbols.

```{r filter_probes}
e2s = toTable(hta20transcriptclusterSYMBOL)
prob_ids = rownames(exprs(eset))
keep_ids = which(prob_ids %in% e2s$probe_id)
eset = ExpressionSet(assayData = exprs(eset)[keep_ids,],
                     phenoData = phenoData(eset),
                     experimentData = experimentData(eset),
                     annotation = annotation(eset))
```

# Data Exploration

PCA plot of normalized expressions.

```{r pca_plot}
res_pca = prcomp(t(exprs(eset)),scale. = FALSE)
screeplot(res_pca)
sample_table = eset@phenoData@data
asp_ratio = res_pca$sdev[2]^2/res_pca$sdev[1]^2
autoplot(res_pca, data = sample_table, colour = 'treatment', 
         shape = 'condition', size = 3, asp = asp_ratio)
sample_names = paste(sample_table$sample_id,
                     sample_table$treatment,sep = "_")
rownames(sample_table) = sample_names
autoplot(res_pca, data = sample_table, 
         shape = FALSE, label = TRUE, label.size = 3, asp = asp_ratio)
```

# Differential Expression Analyses

Use ``limma`` for linear models to assess difference in expression. Define design matrix.

```{r design_matrix}
design = model.matrix(~ time_to_delivery + gestage_delivery, sample_table)
colnames(design) = str_replace(string = colnames(design),
                               pattern = ":",replacement = "_") %>%
  str_replace(string = .,pattern = "treatment",replacement = "")
colnames(design)
```

Automatic independent filtering: 

1. Estimate the correlation between measurements made on the same subject. 
2. Then this inter-subject correlation is input into the linear model fit. 
3. Then compute moderated t-tests.
4. The topTable command provides us a way of ranking genes for further evaluation. In the case below, we adjust for multiple testing by FDR. 

```{r automatic_independent_filtering}
mean_expr = rowMeans(exprs(eset))
thres_candidates = seq(min(mean_expr),quantile(mean_expr, probs = 0.95),1)
fit_list = lapply(thres_candidates, function(thres) {
  cat("Automatic independent filtering: thres = ", thres,"\n")
  # threshold
  eset_thres = ExpressionSet(assayData = exprs(eset)[mean_expr >= thres,],
                             phenoData = phenoData(eset),
                             experimentData = experimentData(eset),
                             annotation = annotation(eset))
  # fit model
  fit = lmFit(eset_thres, design)
  eBayes(fit)
})
num_sig = sapply(fit_list, function(fit) {
  coeffs = colnames(design)[-1]
  gene_table_combined = lapply(coeffs,function(coeff_name) {
    gene_table = topTable(fit, coef = coeff_name, adjust = "BH", 
                          number = nrow(fit))
    gene_table %>% dplyr::filter(adj.P.Val < 0.1) %>% 
      add_column(coeff = coeff_name)
  }) %>% bind_rows()
  nrow(gene_table_combined)
})
num_sig
fit = fit_list[[which.max(num_sig)]]
```

Save results in a list of tables.

```{r top_table}
coeffs = colnames(design)[-1]
gene_table_list = lapply(coeffs, function(coeff_name) {
  gene_table = topTable(fit, 
                        coef = coeff_name, 
                        adjust = "BH", 
                        number = nrow(fit))
  hist(gene_table$P.Value,breaks = 100,main = coeff_name)
  gene_table
})
names(gene_table_list) = coeffs
```

Volcano plots for quality control.

```{r volcanoplot}
for(coeff_name in coeffs) {
  cat(coeff_name,"\n")
  volcanoplot(fit,coef = coeff_name,highlight = 10)
}
```

Map between manufacturer identifiers and gene symbols.

```{r map_probes_to_symbols}
e2s = toTable(hta20transcriptclusterSYMBOL)
map_gene_symbol = function(gene_table) {
  prob_ids = rownames(gene_table)
  symbol = sapply(prob_ids,function(prob_id) {
    matching_symbol = e2s$symbol[prob_id==e2s$probe_id]
    if(length(matching_symbol)==0) matching_symbol = "No_Symbol_Found"
    matching_symbol
  }) %>% unlist
  gene_table = cbind(gene_table,symbol=symbol,stringsAsFactors=FALSE)
  gene_table
}
gene_table_list = lapply(gene_table_list,map_gene_symbol)
```

Print genes that are below an FDR of $0.1$.

```{r print_genes}
for(i in 1:length(gene_table_list)) {
  separator = "-----------"
  cat(separator,names(gene_table_list)[i],separator,"\n")
  gene_table_subset = subset(gene_table_list[[i]],adj.P.Val < 0.1)
  print(gene_table_subset[,c("logFC","adj.P.Val","symbol")])
  cat(separator,names(gene_table_list)[i],separator,"\n\n")
}
```

Write to text file.

```{r write_genes}
for(i in 1:length(gene_table_list)) {
  file_name_results = paste0(names(gene_table_list)[i],"_results.csv")
  cat("writting:",file_name_results,"\n")
  write.csv(gene_table_list[[i]],file = file_name_results)
}
```

# Network Analysis

Network analysis on time to event table.

```{r gene_table_for_network_analysis}
gene_table = gene_table_list[[1]]
```

## Download Gene Network

Download proteins for human species (code is 9606). Consider interaciton that are $0.9$ confidence. From the STRING website: "In STRING, each protein-protein interaction is annotated with one or more 'scores'. Importantly, these scores do not indicate the strength or the specificity of the interaction. Instead, they are indicators of confidence, i.e. how likely STRING judges an interaction to be true, given the available evidence. All scores rank from $0$ to $1$, with $1$ being the highest possible confidence. A score of 0.5 would indicate that roughly every second interaction might be erroneous (i.e., a false positive)."

```{r db_create_string}
string_db = STRINGdb$new(version="10", species = 9606, 
                         score_threshold = 900, input_directory = "")
```

Check how many proteins are in the database.

```{r db_get_proteins}
string_proteins = string_db$get_proteins()
dim(string_proteins)
```

Map gene names to identifiers used in the database.

```{r db_map}
mapped = string_db$map(gene_table, "symbol", removeUnmappedRows = TRUE)
interactions = string_db$get_interactions(mapped$STRING_id)
interactions = data.frame(from = interactions$from,
                          to = interactions$to,
                          combined_score = interactions$combined_score)
dim(interactions)
head(interactions)
```

## Find Subgraph

Fit a Beta-Uniform model.

```{r fit_bum_model, fig.wide=TRUE}
pval = mapped$P.Value
names(pval) = mapped$STRING_id
fb = fitBumModel(pval)
fb
```

Set the `fdr` parameter which can be interpreted as the FDR of the subgraph. Smaller values will produce a smaller maximum subgraph. You should try a few values (e.g. $0.05$, $0.01$, $0.001$) to obtain a reasonable small subgraph that permits biological interpretation. First, we convert the interaction table into an igraph object and make the nodes names human readible. Then we search for the optimal subgraph.

```{r run_fast_heinz}
fdr_vec = seq(0.1,0.01,-0.01)
network = graph_from_data_frame(interactions)
module_list = mclapply(fdr_vec, function(fdr) {
  scores = scoreNodes(network, fb, fdr = fdr)
  module = runFastHeinz(network, scores)
  module
}, mc.cores = 4)
module_lengths = sapply(module_list,function(module) length(V(module)))
tb_module = tibble(id = 1:length(module_lengths), 
                   num_nodes = module_lengths, 
                   fdr = fdr_vec)
ggplot(tb_module, aes(fdr,num_nodes)) +
  geom_line() +
  geom_point(size = 2)
model_sel = tb_module %>% filter(fdr == 0.05) %>% .$id
module = module_list[[model_sel]]
module
```

## Visualize Network

Differential expression is coloured in red (upregulated), green (downregulated), and white (neutral). Shapes represent scores: rectangles are negative and circles are positive.

```{r plot_network, fig.wide=TRUE}
plot_network = function(color_name) {
  if(length(V(module)) == 0) stop("no signficant subgraph found")
  set.seed(0xdada)
  module_df = ggnetwork(module, layout = "kamadakawai")
  module_df$x = c(module_df$x)
  module_df$y = c(module_df$y)
  module_df$xend = c(module_df$xend)
  module_df$yend = c(module_df$yend)
  module_df %<>% mutate(sign_score = factor(sign(module_df$score)))
  ids = sapply(module_df$vertex.names, function(id) which(id == mapped$STRING_id)[1])
  module_df %<>% mutate(t_statistic = mapped$t[ids])
  module_df %<>% mutate(log_fc = mapped$logFC[ids])
  module_df %<>% mutate(symbol = mapped$symbol[ids])
  ggplot(module_df, aes(x = x, y = y, xend = xend, yend = yend)) +
    geom_edges() +
    geom_nodelabel_repel(aes(label = symbol), 
                         box.padding = unit(1, "lines"),
                         alpha = 0.3) +
    geom_nodes(aes_string(shape = "sign_score", color = color_name), size = 6) +
    ggtitle(paste0("Time to Delivery (FDR = ",fdr_vec[model_sel],")")) +
    scale_color_gradient2(midpoint = 0, low = "blue", mid = "white",
                          high = "red", space = "Lab" ) +
    theme_blank()
}
plot_network(color_name= "t_statistic")
ggsave(filename = "network_t_statistic.pdf", width = 10, height = 6)
plot_network("log_fc")
ggsave(filename = "network_log_fc.pdf", width = 10, height = 6)
```

## Single Genes Expression

Scatter plot for some genes in the network.

```{r scatter_plot, fig.width=4, fig.height=4}
plot_node = function(gene_name) {
  probe_id = e2s[which(e2s$symbol == gene_name), "probe_id"]
  tb_expr = tibble(
    expr = exprs(eset)[which(rownames(exprs(eset)) == probe_id),],
    time_to_delivery = pData(eset)$time_to_delivery,
    gestage_delivery = pData(eset)$gestage_delivery
  )
  fit_expr = lm(expr ~ gestage_delivery, tb_expr)
  fit_time = lm(time_to_delivery ~ gestage_delivery, tb_expr)
  tb_expr %<>% mutate(res_expr = residuals(fit_expr))
  tb_expr %<>% mutate(res_time_to_delivery = residuals(fit_time))
  ggplot(tb_expr, aes(res_time_to_delivery, res_expr)) +
    geom_smooth(method = lm, se = FALSE, color = "grey", size = 2) +
    geom_point(size = 2) +
    scale_x_reverse() +
    ggtitle(gene_name) +
    xlab("resid(lm(time_to_delivery ~ gestage_delivery)") +
    ylab("resid(lm(expr ~ gestage_delivery))")
}
plot_node("IFNL1")
plot_node("SNORD18A")
```

# Pathways Analysis

Calculate number of genes that our network on known pathways have in common.

```{r calculate_pathway_overlap}
# get human pathways form kegg database
pathways = keggList("pathway", "hsa")
human_pathways = sub("path:", "", names(pathways))
# kegg server only allow 10 request at the time
n_request = ceiling(length(human_pathways)/10)
chunk = function(x, n) split(x, sort(rank(x) %% n))
chunks_pathways = chunk(1:length(human_pathways),n_request)
# download from kegg server
list_chunk_pathways = lapply(chunks_pathways,function(one_chunk_pathways) {
  cat("download chunk:",one_chunk_pathways,"\n")
  pathway_ids = human_pathways[one_chunk_pathways]
  setNames(keggGet(pathway_ids), pathway_ids)
  })
# flatten list of lists
all_pathways = unlist(list_chunk_pathways, recursive=FALSE)
# check for all human pathways
tb_pways = lapply(all_pathways, function(pway) {
  gene_desc = pway$GENE[c(F,T)]
  if(length(gene_desc) > 0) {
    gene_symbol = sapply(strsplit(gene_desc, split = ";"), 
                         function(desc) desc[1])
    overlap = mean(gene_symbol %in% unique(module_df$symbol))
  } else {
    overlap = 0
  }
  tibble(
    name = pway$NAME,
    overlap = overlap
    )
}) %>% bind_rows
```

Only keep pathways that overlap at least 5% of genes.

```{r filter_pathways}
tb_pways %<>% 
  dplyr::filter(overlap > 0.04) %>% 
  dplyr::arrange(desc(overlap))
tb_pways$name %<>% str_replace(" - Homo sapiens \\(human\\)","")
tb_pways
```

Visualize pathway overlap.

```{r plot_pathways}
tb_pways$name %<>% factor(levels = rev(tb_pways$name))
ggplot(tb_pways, aes(x = 100*overlap,y = name)) + 
  geom_point(size = 2) +
  theme(legend.position="none") +
  xlab("Overlap in %") +
  ylab("Pathway Name")
```

# Session Info {.unnumbered}

```{r session_info}
sessionInfo()
```
